---
title: "1 - Loading in count data, normalization & QC"
author: "Joshua Soon"
date: '25 August 2022'
output: html_document
editor_options: 
  chunk_output_type: console
---



Default dependencies here and yaml are required for config.yml parsing to obtain parameters.
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(yaml)
```



Load the packages for this specific Rmd notebook.
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(DESeq2)
library(edgeR)
library(ggplot2)
```



Load the config yaml file and test a few key parameters.
```{r echo=FALSE, warning=FALSE, message=FALSE}
config <- yaml.load_file("config.yml")

print(config$Paths$Treatment_vs_control_meta)
```



Load the CSV file with the read count data as well as the metadata file containing information about the sample grouping.
The metadata file is needed for DESeq2's grouping.
```{r echo=FALSE, warning=FALSE, message=FALSE}
# csv file must not be UTF-8 format
# Read only if counts have been saved to CSV
RNA_seq_data <- read.csv(data_file)

# For Rh30
# If there are duplicate gene names, take the sum of all read-counts.
# TODO this can be improved upon.
RNA_seq_counts_consolidated <- {RNA_seq_counts_df %>% group_by(Ensembl) %>% summarise_all(sum)}
# Convert the gene symbols to the row names.
RNA_seq_counts_final <- {RNA_seq_counts_consolidated %>% remove_rownames %>% column_to_rownames(var="Ensembl")}

RNA_seq_meta_path <- config$Paths$Treatment_vs_control_meta
RNA_seq_metadata <- read.table(RNA_seq_meta_path, header=T, row.names=1)
```



Check that sample names match in both files
```{r echo=FALSE, warning=FALSE, message=FALSE}
# This must match for the DESeq2 object.
all(colnames(RNA_seq_counts_final) %in% rownames(RNA_seq_metadata))
all(colnames(RNA_seq_counts_final) == rownames(RNA_seq_metadata))
```



Plot RNA-seq count distribution for all samples
```{r echo=FALSE, warning=FALSE, message=FALSE}
# This can be done for other samples as well for QC-ing.
ggplot(RNA_seq_counts_final) +
  xlim(-5, 100) +
  geom_histogram(aes(x = Gene_xyz_KO_1), stat = "bin", bins = 100) +
  xlab("Raw expression counts") +
  ylab("Number of genes")

ggplot(RNA_seq_counts_final) +
  xlim(-5, 100) +
  geom_histogram(aes(x = Gene_xyz_KO_2), stat = "bin", bins = 100) +
  xlab("Raw expression counts") +
  ylab("Number of genes")

ggplot(RNA_seq_counts_final) +
  xlim(-5, 100) +
  geom_histogram(aes(x = Gene_xyz_KO_3), stat = "bin", bins = 100) +
  xlab("Raw expression counts") +
  ylab("Number of genes")

ggplot(RNA_seq_counts_final) +
  xlim(-5, 100) +
  geom_histogram(aes(x = Gene_xyz_KO_4), stat = "bin", bins = 100) +
  xlab("Raw expression counts") +
  ylab("Number of genes")

ggplot(RNA_seq_counts_final) +
  xlim(-5, 100) +
  geom_histogram(aes(x = Wildtype_1), stat = "bin", bins = 100) +
  xlab("Raw expression counts") +
  ylab("Number of genes")

ggplot(RNA_seq_counts_final) +
  xlim(-5, 100) +
  geom_histogram(aes(x = Wildtype_2), stat = "bin", bins = 100) +
  xlab("Raw expression counts") +
  ylab("Number of genes")

ggplot(RNA_seq_counts_final) +
  xlim(-5, 100) +
  geom_histogram(aes(x = Wildtype_3), stat = "bin", bins = 100) +
  xlab("Raw expression counts") +
  ylab("Number of genes")

ggplot(RNA_seq_counts_final) +
  xlim(-5, 100) +
  geom_histogram(aes(x = Wildtype_4), stat = "bin", bins = 100) +
  xlab("Raw expression counts") +
  ylab("Number of genes")
```



Modeling count data for Treatment and Control conditions.
For RNA-seq with 4 treatment and 4 control, treatment = columns 1:4, treatment = columns 5:8
```{r echo=FALSE, warning=FALSE, message=FALSE}
treatment_mean_counts <- apply(RNA_seq_counts_final[, 1:4], 1, mean) # Control means
treatment_variance_counts <- apply(RNA_seq_counts_final[, 1:4], 1, var) # Control Variance
treatment_variance_vs_means <- data.frame(treatment_mean_counts, treatment_variance_counts)

ggplot(treatment_variance_vs_means) +
  geom_point(aes(x=treatment_mean_counts, y=treatment_variance_counts)) + 
  geom_line(aes(x=treatment_mean_counts, y=treatment_mean_counts, color="red")) +
  guides(color=guide_legend(title="Treatment\nvariance = mean")) +
  scale_y_log10() +
  scale_x_log10()

control_mean_counts <- apply(RNA_seq_counts_final[, 5:8], 1, mean) # Treatment means
control_variance_counts <- apply(RNA_seq_counts_final[, 5:8], 1, var) # Treatment variance
control_variance_vs_means <- data.frame(control_mean_counts, control_variance_counts)

ggplot(control_variance_vs_means) +
  geom_point(aes(x=control_mean_counts, y=control_variance_counts)) + 
  geom_line(aes(x=control_mean_counts, y=control_mean_counts, color="red")) +
  guides(color=guide_legend(title="Control\nvariance = mean")) +
  scale_y_log10() +
  scale_x_log10()
```



###################
## Optional part ##
###################

Filter out low read-counts before creating DESeq2Dataset object
This is done with edgeR's cpm function.
```{r echo=FALSE, warning=FALSE, message=FALSE}
RNA_seq_edgeR_DEG <- DGEList(RNA_seq_counts_final)

# Calculate normalization factors
RNA_seq_edgeR_DEG <- calcNormFactors(RNA_seq_edgeR_DEG)

# Filter low-expressed genes, less than 0.2 cpm (counts per million)
cpm_cutoff <- 0.2

# Min cpm of each row of cpm(gene_counts), that is less than cutoff
genes_to_drop <- which(apply(cpm(RNA_seq_edgeR_DEG), 1, min) < cpm_cutoff)
RNA_seq_cpm_filtered <- RNA_seq_counts_final[-genes_to_drop,]
```



Actual DESeq2 part

This is the part which is slightly debatable and probably needs one trial and error to see which works best.

On some biostars forum posts, it is not recommended to do any form of CPM filtering before DESeq2 because it automatically
does some form internal filtering before fitting its model. For edgeR, this filtering is needed as it is well described in
their vignette. 

So both options to add CPM filtering or use the raw counts straight into DESeq2 can be tried. For this, the variable "RNA_seq_counts_final" can be swapped for "RNA_seq_cpm_filtered" to use the CPM filtered one instead.

DESeq2 object
```{r echo=FALSE, warning=FALSE, message=FALSE}
### Create DESeq2Dataset object
RNA_seq_dds <- DESeqDataSetFromMatrix(countData = round(RNA_seq_counts_final), colData = RNA_seq_metadata, design = ~ sampletype)
View(counts(RNA_seq_dds))
RNA_seq_dds <- estimateSizeFactors(RNA_seq_dds)
sizeFactors(RNA_seq_dds)
RNA_seq_normalized_counts <- counts(RNA_seq_dds, normalized=TRUE)
data_name <- "Treatment_vs_control"
normalized <- "_normalized_counts.txt"
write.table(RNA_seq_normalized_counts, file = paste(data_name, normalized, sep=""), sep="\t", quote=F, col.names=NA)

### Total number of raw counts per sample
total_raw_counts <- as.data.frame(colSums(counts(RNA_seq_dds)))

## Total number of normalized counts per sample
total_normalized_counts <- colSums(counts(RNA_seq_dds, normalized=T))
```
