---
title: "5 - Post DESeq2 filtering and result annotation"
author: "Joshua Soon"
date: '28 August 2022'
output: html_document
editor_options: 
  chunk_output_type: console
---

This additional step post DESeq2 differential expression is mainly to accomplish these few objectives:
1. Filtering out genes with less than a CPM of 0.2
2. Annotate the filtered results with HGCN gene symbols, Entrez IDs & gene descriptions
3. Other filtering based on Log2FoldChange or Adjust p-value.
4. Export CPM edgeR normalized counts after CPM > 0.2 filter for GSEA.



Default dependencies here and yaml are required for config.yml parsing to obtain parameters.
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(yaml)
```



Load the packages for this specific Rmd notebook.
```{r}
library(org.Hs.eg.db)
library(tidyverse)
library(dplyr)
library(edgeR)
```

```{r}
source(here::here("src", "gsea_helpers.R"))
```



Drop those genes which fail to meet the low CPM cutoff.
RNA-seq differential expression results (DE) with and without low CPM filtering.
```{r echo=FALSE, warning=FALSE, message=FALSE}
RNA_seq_DE_cpm_filtered <- subset(RNA_seq_DE_table_rn, !(rownames(RNA_seq_DE_table_rn) %in% names(genes_to_drop)))
```



Annotate the DESeq2 results with Gene symbols, Entrez IDs and brief gene descriptions
```{r echo=FALSE, warning=FALSE, message=FALSE}
# For CPM filtered DE results:
res_annotations <- AnnotationDbi::select(org.Hs.eg.db,
                                         keys = rownames(RNA_seq_DE_cpm_filtered),
                                         columns = c ("SYMBOL", "ENTREZID", "GENENAME"),
                                         keytype = "ENSEMBL") 

RNA_seq_DE_cpm_filtered_annotated <- merge(RNA_seq_DE_cpm_filtered, res_annotations, by.x=0, by.y="ENSEMBL")

# Save outputs to a CSV file
write.csv(RNA_seq_DE_cpm_filtered_annotated, "./RNA_seq_differential_expression_cpm_filtered.csv", row.names = FALSE)

# For non filtered DE results:
all_res_annotations <- AnnotationDbi::select(org.Hs.eg.db,
                                             keys = rownames(RNA_seq_DE_table_rn),
                                             columns = c ("SYMBOL", "ENTREZID", "GENENAME"),
                                             keytype = "ENSEMBL") 

RNA_seq_DE_annotated <- merge(RNA_seq_DE_table_rn, all_res_annotations, by.x=0, by.y="ENSEMBL")

write.csv(RNA_seq_DE_annotated, "./RNA_seq_differential_expression.csv", row.names = FALSE)
```



This section might be somewhat similar to the sigDEGs portion in the 3rd Rmd, but this code block would save the significant
differentially expressed genes along with the gene annotations.

Only run the write.csv() functions to export the CSVs for the annotated significant differentially expressed genes.
```{r echo=FALSE, warning=FALSE, message=FALSE}
###################################################
### Do not commit the outputs from here to Git! ###
###################################################

padj_cutoff <- 0.05
l2fc_cutoff <- 1

# Only retain genes that are above a 0.2 CPM, below 0.05 for their adjusted p-value
# and are differentially expressed greater than a Fold Change of 2 (L2FC = 1).
RNA_seq_DE_annotated_sigDEGS <- RNA_seq_DE_annotated %>% dplyr::filter(padj < padj_cutoff & abs(log2FoldChange) > l2fc_cutoff)
View(RNA_seq_DE_annotated_sigDEGS)

RNA_seq_DE_annotated_sigDEGS_cpm_filtered <- RNA_seq_DE_cpm_filtered_annotated %>% dplyr::filter(
  padj < padj_cutoff 
  & abs(log2FoldChange) > l2fc_cutoff
  )
View(RNA_seq_DE_annotated_sigDEGS_cpm_filtered)

write.csv(RNA_seq_DE_annotated_sigDEGS, "./RNA_seq_annotated_sigDEGS.csv", row.names = FALSE)
```



edge R normalization for export to GSEA
```{r}
# Normal edge R normalization
limma_voom_normalized <- cpm(RNA_seq_cpm_filtered, log=F)
# Normalization with log
limma_voom_normalized_log <- cpm(RNA_seq_cpm_filtered, log=TRUE, prior.count=1)

# Convert results to dataframe
limma_voom_normalized_df <- data.frame(limma_voom_normalized)
limma_voom_normalized_log_df <- data.frame(limma_voom_normalized_log)

# Save as CSVs
write.csv(limma_voom_normalized_df, "./edgeR_normalized_for_GSEA.csv", row.names = T)
write.csv(limma_voom_normalized_log_df, "./edgeR_log_normalized_for_GSEA.csv", row.names = T)
```
