---
title: "6 - Functional enrichment analysis (cluster profiler)"
author: "Joshua Soon"
date: '28 August 2022'
output: html_document
editor_options: 
  chunk_output_type: console
---



Load the packages for this specific Rmd notebook.
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(clusterProfiler)
library(DOSE)
library(ReactomePA)
library(enrichplot)
library(ggnewscale)
library(ggplot2)
library(org.Hs.eg.db)
library(msigdbr)
```



The genes extracted for Functional Enrichment will be named as `FE_genes_` and suffixed with other relevant identifiers, like with and without CPM filtering.
For functional enrichment, we are most interested in the significant differently expressed genes:
1. RNA_seq_DE_annotated_sigDEGS (without low CPM filtering)
2. RNA_seq_DE_annotated_sigDEGS_cpm_filtered (with low CPM filtering)
For this analysis, we will proceed with the genes with post DESeq2 CPM filtering but the other option is possible too.

For various cutoffs on Log2FoldChange and adjusted p-value, the changes can be made in 5 - Post DESeq2 filtering and result annotation.

Within the results, the HGNC gene symbols and Entrez IDs for enrichKEGG would be found in the SYMBOL & ENTREZID columns respectively.
```{r echo=FALSE, warning=FALSE, message=FALSE}
FE_genes_cpm_filtered <- RNA_seq_DE_annotated_sigDEGS_cpm_filtered

# The Row.names column contain the Ensembl IDs
FE_genes_ensembl_all <- FE_genes_cpm_filtered$Row.names
length(FE_genes_ensembl_all)
# Get unique ones
FE_genes_ensembl_unique <- unique(FE_genes_ensembl_all)

# The ENTREZID column will contain the Entrez IDs
FE_genes_entrez_all <- FE_genes_cpm_filtered$ENTREZID
sum(is.na(FE_genes_entrez_all))

# Remove NAs for those which cannot map to an Entrez IDs
entrez_no_na <- na.omit(FE_genes_entrez_all)
length(entrez_no_na)

# Get unique ones
FE_genes_entrez_unique <- unique(entrez_no_na)
length(FE_genes_entrez_unique)
```



## Optional
Performing functional enrichment on up- and down-regulated genes separately.
In the event which the user wishes to perform functional enrichment on up- and down-regulated genes separately, this cell outputs the necessary variables for that.
The unique Ensembl and Entrez IDs can be retrieve from the up_reg and down_reg dataframes individually and be piped into the downstream functional enrichment steps.
`FE_genes_up_reg_ensembl_unique` and `FE_genes_up_reg_entrez_unique` would have the Ensembl and Entrez IDs for up-regulated genes respectively.
`FE_genes_down_reg_ensembl_unique` and `FE_genes_down_reg_entrez_unique` would have the Ensembl and Entrez IDs for down-regulated genes respectively.
```{r}
FE_genes_up_reg_cpm_filtered <- RNA_seq_DE_cpm_filtered_annotated %>% dplyr::filter(
  padj < padj_cutoff 
  & log2FoldChange > 1
  )

# The Row.names column contain the Ensembl IDs
FE_genes_up_reg_ensembl_all <- FE_genes_up_reg_cpm_filtered$Row.names
length(FE_genes_up_reg_ensembl_all)
# Get unique ones
FE_genes_up_reg_ensembl_unique <- unique(FE_genes_up_reg_ensembl_all)

# The ENTREZID column will contain the Entrez IDs
FE_genes_up_reg_entrez_all <- FE_genes_up_reg_cpm_filtered$ENTREZID
sum(is.na(FE_genes_up_reg_entrez_all))

# Remove NAs for those which cannot map to an Entrez IDs
entrez_no_na <- na.omit(FE_genes_up_reg_entrez_all)
length(entrez_no_na)

# Get unique ones
FE_genes_up_reg_entrez_unique <- unique(entrez_no_na)
length(FE_genes_up_reg_entrez_unique)

FE_genes_down_reg_cpm_filtered <- RNA_seq_DE_cpm_filtered_annotated %>% dplyr::filter(
  padj < padj_cutoff 
  & log2FoldChange < -1
  )

# The Row.names column contain the Ensembl IDs
FE_genes_down_reg_ensembl_all <- FE_genes_down_reg_cpm_filtered$Row.names
length(FE_genes_down_reg_ensembl_all)
# Get unique ones
FE_genes_down_reg_ensembl_unique <- unique(FE_genes_down_reg_ensembl_all)

# The ENTREZID column will contain the Entrez IDs
FE_genes_down_reg_entrez_all <- FE_genes_down_reg_cpm_filtered$ENTREZID
sum(is.na(FE_genes_down_reg_entrez_all))

# Remove NAs for those which cannot map to an Entrez IDs
entrez_no_na <- na.omit(FE_genes_down_reg_entrez_all)
length(entrez_no_na)

# Get unique ones
FE_genes_down_reg_entrez_unique <- unique(entrez_no_na)
length(FE_genes_down_reg_entrez_unique)
```



Gene Ontology (GO) over representation analysis (ORA)
1. Biological Processes (BP)
2. Molecular Function (MF)
3. Cellular Component (CC)

GO BP
```{r echo=FALSE, warning=FALSE, message=FALSE}
GO_BP <- enrichGO(gene = FE_genes_ensembl_unique,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "BP",
                  pAdjustMethod = "BH",
                  qvalueCutoff = 0.05,
                  readable = TRUE)

GO_BP_bar_plot <- barplot(GO_BP, showCategory=10, font.size = 16) +
  xlab("Number of genes") +
  ggtitle("Barplot for EnrichGO:BP")

GO_BP_bar_plot

GO_BP_dot_plot <- dotplot(GO_BP, showCategory=10, font.size = 16) +
  ggtitle("Dotplot for EnrichGO:BP")

GO_BP_dot_plot
```

GO MF
```{r echo=FALSE, warning=FALSE, message=FALSE}
GO_MF <- enrichGO(gene = FE_genes_ensembl_unique,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "MF",
                  pAdjustMethod = "BH",
                  qvalueCutoff = 0.05,
                  readable = TRUE)

GO_MF_bar_plot <- barplot(GO_MF, showCategory=10, font.size = 16) +
  xlab("Number of genes") +
  ggtitle("Barplot for EnrichGO:MF")

GO_MF_bar_plot

GO_MF_dot_plot <- dotplot(GO_MF, showCategory=10, font.size = 16) +
  ggtitle("Dotplot for EnrichGO:MF")

GO_MF_dot_plot
```

GO CC
```{r echo=FALSE, warning=FALSE, message=FALSE}
GO_CC <- enrichGO(gene = FE_genes_ensembl_unique,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "CC",
                  pAdjustMethod = "BH",
                  qvalueCutoff = 0.05,
                  readable = TRUE)

GO_CC_bar_plot <- barplot(GO_CC, showCategory=10, font.size = 16) +
  xlab("Number of genes") +
  ggtitle("Barplot for EnrichGO:CC")

GO_CC_bar_plot

GO_CC_dot_plot <- dotplot(GO_CC, showCategory=10, font.size = 16) +
  ggtitle("Dotplot for EnrichGO:CC")

GO_CC_dot_plot
```



Over-representation analysis for disease-gene associations.
For this, Entrez IDs are needed so `FE_genes_entrez_unique` is used.
```{r echo=FALSE, warning=FALSE, message=FALSE}
DGN <- enrichDGN(gene = FE_genes_entrez_unique, pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)

DGN_dot_plot <- dotplot(DGN, showCategory=10, font.size = 16) + ggtitle("Dotplot for EnrichDGN")

DGN_dot_plot
```



Over-representation analysis for the Network of Cancer Genes (NCG).
```{r echo=FALSE, warning=FALSE, message=FALSE}
NCG <- enrichNCG(gene = FE_genes_entrez_unique, pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)

NCG_dot_plot <- dotplot(NCG, showCategory=10, font.size = 16) + ggtitle("Dotplot for EnrichNCG")

NCG_dot_plot
```



Disease ontology over-representation analysis.
```{r echo=FALSE, warning=FALSE, message=FALSE}
DO <- enrichDO(gene = FE_genes_entrez_unique, pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)

DO_dot_plot <- dotplot(DO, showCategory=10) + ggtitle("Dotplot for EnrichDO")

DO_dot_plot
```



Reactome pathway over-representation analysis.
```{r echo=FALSE, warning=FALSE, message=FALSE}
Reactome_PA <- enrichPathway(gene = FE_genes_entrez_unique, pAdjustMethod = "BH", qvalueCutoff = 0.05, readable=TRUE)

Reactome_PA_dot_plot <- dotplot(Reactome_PA, showCategory=10) + ggtitle("Dotplot for Reactome pathway")

Reactome_PA_dot_plot
```



KEGG pathway over-representation analysis.
```{r}
KEGG <- enrichKEGG(gene = FE_genes_entrez_unique, pAdjustMethod = "BH", qvalueCutoff = 0.05)

KEGG_dot_plot <- dotplot(KEGG, showCategory=10) + ggtitle("Dotplot for EnrichKEGG")

KEGG_dot_plot
```



Enrichment maps
GO BP enrichment map
```{r}
GO_BP_pairwise <- pairwise_termsim(GO_BP)

GO_BP_enrichment_map <- emapplot(GO_BP_pairwise, showCategory = 30, min_edge = 0.2, cex_label_category = 1)

GO_BP_enrichment_map
```

GO MF enrichment map
```{r}
GO_MF_pairwise <- pairwise_termsim(GO_MF)

GO_MF_enrichment_map <- emapplot(GO_MF_pairwise, showCategory = 30, min_edge = 0.2, cex_label_category = 1)

GO_MF_enrichment_map
```

GO CC enrichment map
```{r}
GO_CC_pairwise <- pairwise_termsim(GO_CC)

GO_CC_enrichment_map <- emapplot(GO_CC_pairwise, showCategory = 30, min_edge = 0.2, cex_label_category = 1)

GO_CC_enrichment_map
```
